Staged Abstract Interpreters

### 2019-08-11 paper/10_intro.tex
revise intro

The central basis of our approach is a semantic-agnostic interpreter, which can be
instantiated by different interpretations or compilations (Figure~\ref{confluence}).
The different semantics in Figure~\ref{confluence} correspond to four different
artifacts involved in software language engineering: definitional interpreters
which are defined by the denotational or operational semantics, compilers that
translate the program into another language, analyzers that statically compute the
runtime properties of programs, and optimizing analyzers that achieve the same goal
of analyzers but in the way of compilers. We observe that these four
tasks may look like very differently at first glance, but in fact are
all firmly rooted in the concrete semantics of the language.

To achieve the flexibility of reinterpreting the shared interpreter by
different semantics, our framework abstracts the primitive operations
from these semantics, and extensively uses abstract type members over
the binding-times, the components of the interpreter, and the returned
monadic type of the interpreter. The binding-time abstraction is
represented by a higher-kinded type, and controls whether the
interpreter produces values or generates code. The components of the
interpreter, such as the environment and values, determines the domain
that the interpreter will be running with. Following the AAM approach,
by applying a series of abstractions to these components, we can
derive computable and sound abstract interpreters. Finally, the
monadic abstraction encapsulates both the binding-times and
components, which appears as the returned type and defines the
behaviors and datatype the interpreter operates on.

\paragraph{Applications and Evaluations}
We evaluate the idea of staging an abstract interpreter through
multiple case studies and empirical evaluation on performance.
1) We compare our approach with abstract compilation
\cite{Boucher:1996:ACN:647473.727587}, and show that we can achieve
the same optimization by utilizing type-based stage annotations;
meanwhile, the analyzer program does not need any changes, thereby
requires little engineering efforts.
2) We extend the basic staged abstract interpreter to diverse flow
analysis, including a store-widened analysis, a context-sensitive
analysis and abstract garbage collection.
3) We notice that staging an abstract interpreter enables modularly
compiling an analysis to programs. The compiled analysis is reusable,
therefore it provides a mechanized way to create library summaries by
reusing a whole-program analyzer.
4) We also empirically evaluate the performance improved by staging,
showing an order of magnitude speed-up on flow-analysis.

>>>>

Program specialization and abstract interpretation are two orthogonal concepts.
To implement the confluence of them, we first construct a generic interpreter that
is agnostic to both binding-times and value domains used in the semantics.
Later, the generic interpreter can be instantiated from these two dimensions (Figure~\ref{confluence}):
\begin{itemize}
\item With a flat binding-time and concrete domains, it is a ordinary definitional interpreter based
  on big-step operational semantics;
\item With two-level binding-times and concrete domain, it is a compiler that translate to program
  intro antoher language;
\item With a flat binding-time and abstract domains, it is an definitional abstract interpreter
  \cite{DBLP:journals/pacmpl/DaraisLNH17} that statically computes runtime properties;
\item With two-level binding-times and abstract domains, it is an optimizing program analyzer but
  works in the fashion of compilation.
\end{itemize}

Although the four artifacts may look like very differently at first glance,
but in fact are all firmly rooted in the concrete semantics of the language.
This observation provides a way to abstract over the interpreter and achieve
the flexibility of reinterpreting the shared interpreter. We adapt the monadic
interpreter schema, which already have been applied to abstract interpreters,
and introduce the binding-time abstraction into the generic interpreter.
The generic interpreter returns a value of monadic type, which can be varied
by different semantics. The domain of the interpreter and the effects such as read,
state and nondeterminism can be both wrapped into this monadic type.
The binding-time abstraction is represented by a higher-kinded type,
and controls whether the interpreter produces values directly or generates code.
It is worth to mention that the binding-time type is also instrumented into
the monadic type, so that we will distinguish normal monads and staged monads.

\paragraph{Applications and Evaluations}
We evaluate the idea of staging an abstract interpreter through
case studies and an empirical evaluation on performance.
1) We compare our approach with abstract compilation
\cite{Boucher:1996:ACN:647473.727587}, a implementation technique for
control-flow analyses, and show that by utilizing type-based stage
annotations we can achieve the same optimization, and meanwhile,
the analyzer does not need to change, thereby requires much less
engineering efforts.
2) We extend the basic staged abstract interpreter to different flow
analyses, including a store-widened analysis, a context-sensitive
analysis and abstract garbage collection.
3) We notice that staging an abstract interpreter enables modularly
compiling an analysis to programs. Here we borrow the concept of modular
analysis, and show that the compiled analysis is reusable.
Therefore the approach provides a modular way to create optimized analysis
code by mechanized reusing a whole-program analyzer.
4) We also empirically evaluate the performance improved by staging,
showing an order of magnitude speed-up on flow-analysis \todo{explicit note the number}.
